---
title: "TLS_FvFm_methods_exp1"
author: "Pieter Arnold"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Libraries

Load in R packages and functions required for the models to run.

```{r}

pacman::p_load(ggplot2, ggpubr, patchwork, scales, tidyr, dplyr, lmerTest, plotrix, readr, qs2, foreach, doParallel, pbapply)

source("Code/TLS_functions.R")

```

## Load data

Import the data to be extracted. Calculate the proportional Fv/Fm based on the time periods of measurements.
Typically, this is initial, 2h post-heat, and 16-24h post-heat. 
Rename columns to a consistent naming convention for the pipeline to run as intended.
At minimum you'll need: Temp, Time, final_fvfm, initial_fvfm


```{r}

# Add filepath
Exp1_wide <- read.csv("Data/Exp1_wide.csv", stringsAsFactors = TRUE)

str(Exp1_wide)

levels(Exp1_wide$Species) <- c("Vicia faba", 
                               "Eucalyptus pauciflora", 
                               "Lomandra longifolia", 
                               "Populus nigra")
Exp1_wide$Species <- factor(Exp1_wide$Species, levels = c("Eucalyptus pauciflora", 
                                                          "Lomandra longifolia", 
                                                          "Populus nigra", 
                                                          "Vicia faba"))

dat <- Exp1_wide
rm(Exp1_wide)

# initial (before heat)      [i]
# overnight (24h after heat) [o]

# Calculate Fv/Fm as proportion (initial - overnight)
dat$fvfm_prop_io <- dat$final/dat$initial
dat$fvfm_prop_io <- ifelse(dat$final/dat$initial=="NaN", 0, dat$final/dat$initial) 
dat$fvfm_prop_io <- ifelse(dat$fvfm_prop_io > 1, 1, dat$fvfm_prop_io) 

# dat <- dat %>% rename(Time = Duration)
dat <- dat %>% rename(final_fvfm = final)
dat <- dat %>% rename(initial_fvfm = initial)
dat$unique_sample <- paste0(dat$AOI)
dat$unique_sample <- as.integer(as.factor(dat$unique_sample))
dat$Species_sample <- paste0(dat$Species, "_", dat$unique_sample)

nlevels(as.factor(unique(dat$unique_sample)))

```


## Clean data

Double check the number of samples, ensuring that most of them have representation across the treatments. Exclude samples that showed damage under benign conditions (clearly unhealthy or physically damaged, unrelated to the heat treatment). Option to manually remove samples that are known to be poor quality.


```{r}

# Number of unique samples 
nsamp <- length(unique(dat$unique_sample))
print(nsamp)

# Check the data depth: length_sample would ideally be > 20
length_sample <- matrix(NA, nsamp, 1)
for (i in 1:nsamp) {
  length_sample[i] <- length(which(dat$unique_sample==i))
}
# Check which sample(s) has poor replication
which(length_sample < 20) # integer(0) indicates all are ok

dat_clean <- dat
dim(dat_clean)
head(dat_clean)

# Create directory for outputs if it doesn't exist
if(!dir.exists("Output")) {
  dir.create("Output")
}

# Save cleaned data
write.csv(dat_clean, "Output/Exp1_dat_clean.csv")

```


## Data summary

Summarise and plot the raw Fv/Fm data of the cleaned dataset overall. This should give insight into ubiquitously problematic treatments from the heat assays. If used, The simulated anchor is included in the plotting.

```{r}


fvfm_prop_summary <- dat_clean %>%
  group_by(Temp, Time) %>%
  summarise(
    meanFvFm = mean(fvfm_prop_io, na.rm = TRUE),
    sd = sd(fvfm_prop_io, na.rm = TRUE),
    se = std.error(fvfm_prop_io, na.rm = TRUE))
fvfm_prop_summary$Temp <- as.factor(fvfm_prop_summary$Temp)
fvfm_prop_summary


ggplot(data = fvfm_prop_summary, aes(x = log10(Time), y = meanFvFm, colour = Temp)) +
  geom_point(position = position_jitter(0.00), alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, fullrange = TRUE) +
  scale_x_continuous(breaks = c(log10(1), log10(5), log10(15), 
                                log10(30), log10(60), log10(120), log10(240)),
                     labels = c("1", "5", "15", "30", "60", "120", "240")) +
  scale_y_continuous(limits = c(-0.03,1)) +
  labs(y = bquote(Proportional~italic(F)[V]/italic(F)[M]), x = bquote(Time~(min)), colour = "Temperature") +
  scale_colour_viridis_d(option = "inferno", end = 0.85) +
  theme_classic()

fvfm_summary <- dat_clean %>%
  group_by(Temp, Time) %>%
  summarise(
    meanFvFm = mean(final_fvfm, na.rm = TRUE),
    sd = sd(final_fvfm, na.rm = TRUE),
    se = std.error(final_fvfm, na.rm = TRUE))
fvfm_summary$Temp <- as.factor(fvfm_summary$Temp)
fvfm_summary

ggplot(data = fvfm_summary, aes(x = log10(Time), y = meanFvFm, colour = Temp)) +
  geom_point(position = position_jitter(0.00), alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, fullrange = TRUE) +
  scale_x_continuous(breaks = c(log10(1), log10(5), log10(15), 
                                log10(30), log10(60), log10(120), log10(240)),
                     labels = c("1", "5", "15", "30", "60", "120", "240")) +
  scale_y_continuous(limits = c(-0.03,0.8)) +
  labs(y = bquote(italic(F)[V]/italic(F)[M]), x = bquote(Time~(min)), colour = "Temperature") +
  scale_colour_viridis_d(option = "inferno", end = 0.85) +
  theme_classic()

```

## User config

User defined parameters for fitting the TLS models
`rep` = number of bootstrap iterations, will increase model fitting time proportionally
`response_vars` = a list of response variables to be calculated. fvfm_io is initial to overnight and is the minimum used. Customisable if additional measurement time points are included.
`thresholds` = a list of thresholds to include in the analysis. T50 is typical (50% decline in Fv/Fm), but also the a point of damage initiation (e.g. T5, T10, T15, 'Tcrit') and a point of substantial damage (e.g. T90, T95). Both the threshold name and value from 0 to 1 need to be defined and matched.

`response_vars_absolute` = a list of response variables to be calculated. fvfm_absolute is the final Fv/Fm value (absolute raw data). Customisable for different thresholds if additional measurement time points are included.
`thresholds` = a list of thresholds for absolute values to include in the analysis. T0.3 is typical (temperature when Fv/Fm = 0.3), but others can be added between the bounds of 0 and ~0.8.

```{r}

# User-defined parameters
# Configuration
config <- list(
  rep = 1000, # Bootstrap replications,  # CHANGE AS REQUIRED
  uniquesamples = length(unique(dat$unique_sample)),
  time_levels = c(5,15,30,60,120),  # CHANGE AS REQUIRED
  resample = TRUE, # Option for bootstrap or original data (recommend using bootstrapping)
  
  # Define which response variables to analyse  # CHANGE AS REQUIRED
  response_vars <- list(
    fvfm_io = list(
      name = "fvfm_prop_io",
      description = "Initial to 24h recovery", 
      thresholds = list(
        T90 = 0.1, # 90% decline
        T50 = 0.5, # 50% decline
        T10 = 0.9  # 10% decline, may not always work if there aren't a few control treatments (low temp, short time)
      )
    )
  ), 
  
  response_vars_absolute <- list(
    fvfm_absolute = list(
      name = "final_fvfm",  # Use the raw Fv/Fm column instead of proportional
      description = "Absolute Fv/Fm values",
      thresholds = list(
        T0.3 = 0.3    # Temperature where Fv/Fm = 0.3
      )
    )
  )
)

config$response_vars <- c(response_vars, response_vars_absolute)

config

```


## Extract data

This applies many of the functions to extract the threshold temperature values under each treatment and then calculate TLS metrics from those fits. Also generates plots for diagnostics. Stores results in `all_results` as a list that is compiled in the next steps. Progress is indicated.

```{r}

# Initialise output lists
all_results <- list()
all_diagnostic_plots <- list()

for(k in 1:config$uniquesamples) {
  tryCatch({
    cat(sprintf("\rProcessing sample %d/%d", k, config$uniquesamples))
    cat("\n")  
    
    # Prepare data
    sp <- dat_clean %>%
      filter(unique_sample == k) %>%
      mutate(logTime = log10(Time))
    
    if(nrow(sp) < 10) {
      cat(sprintf(" - Skipped (insufficient data)\n"))
      next
    }
    
    # Generate prediction grid
    pred_grid <- generate_pred_grid(sp)
    
    if(isTRUE(config$resample)) {
      # Bootstrap predictions for all response variables
      cat("Running with resampling (bootstrap)\n")
      predictions <- bootstrap_predictions(sp, pred_grid, config$response_vars, config$rep, 
                                           resample = TRUE, progress = TRUE, n_cores = parallel::detectCores())
    } else {
      cat("Running without resampling (original)\n")
      predictions <- bootstrap_predictions(sp, pred_grid, config$response_vars, config$rep, 
                                           resample = FALSE, progress = FALSE)
    }
    
    # Calculate temperature thresholds for all responses
    threshold_results <- calculate_all_thresholds(
      predictions, pred_grid, config$response_vars, config$rep
    )
    
    # Calculate TLS metrics for all responses
    tls_metrics <- list()
    for(resp_var in names(config$response_vars)) {
      tls_metrics[[resp_var]] <- calculate_tls_metrics(threshold_results[[resp_var]], verbose = TRUE, 
                                                       min_r_squared = 0.1, 
                                                       temp_range = c(15,80),
                                                       z_range = c(-30,5),
                                                       min_temp_span = 0.1)
      
      # Add metadata
      for(thresh_name in names(tls_metrics[[resp_var]])) {
        if(!is.null(tls_metrics[[resp_var]][[thresh_name]])) {
          tls_metrics[[resp_var]][[thresh_name]]$unique_sample <- k
          tls_metrics[[resp_var]][[thresh_name]]$response_type <- resp_var
        }
      }
    }
    
    # Generate diagnostic plots
    plots <- generate_diagnostic_plots(
      sp, predictions, pred_grid, 
      threshold_results, tls_metrics, 
      k, config$response_vars
    )
    
    # Save plots for each response type
    plot_combined_list <- list()
    
    for(resp_var in names(plots)) {
      resp_name <- gsub("fvfm_", "", resp_var)  # Clean name for filename
      
      # Combine plots for this response
      plot_combined_list[[resp_var]] <- ggarrange(
        plotlist = plots[[resp_var]], 
        nrow = 3, ncol = 2, 
        align = "h", labels = "AUTO"
      )
      
      # Save plots
      for(resp_var in names(plot_combined_list)) {
        resp_name <- gsub("fvfm_", "", resp_var)
        plot_name <- sprintf("sample_%03d_%s", k, resp_name)
        all_diagnostic_plots[[plot_name]] <- plot_combined_list[[resp_var]]
      }
    }
    
    # Store results
    all_results[[k]] <- list(
      thresholds = threshold_results,
      tls_metrics = tls_metrics,
      predictions = predictions
    )
    
  }, error = function(e) {
    cat(sprintf("\nERROR in sample %d: %s\n", k, conditionMessage(e)))
  })
}

pdf("Output/Exp1_all_samples_diagnostics.pdf", width = 8, height = 12)
for(plot in all_diagnostic_plots) {
  print(plot)
}
dev.off()
rm(plots)

```

## Save results

```{r, eval=FALSE}

qs_save(all_results, "Output/Exp1_results_data.qs2", nthreads = parallel::detectCores())  # Small file
# qs_save(all_diagnostic_plots, "Output/results_plots.qs2", nthreads = parallel::detectCores())   # Large file

```


```{r, eval=FALSE}

# Load output list with plots as a .rds file (may read a large file if rep is large)
# all_results <- qs_read("Output/Exp1_results_data.qs2")
# all_diagnostic_plots <- qs_read("Output/Exp1_results_plots.qs2")

```


## Summarize results

```{r}

# Extract all results for each response type
results_summary <- list()

for(resp_var in names(config$response_vars)) {
  results_summary[[resp_var]] <- list()
  
  # Get thresholds
  for(thresh in names(config$response_vars[[resp_var]]$thresholds)) {
    results_summary[[resp_var]]$thresholds[[thresh]] <- 
      compile_results_by_response(all_results, resp_var, thresh)
    
    results_summary[[resp_var]]$tls[[thresh]] <- 
      compile_tls_by_response(all_results, resp_var, thresh)
  }
}

```

## Save results

```{r}

for(resp in names(results_summary)) {
  for(thresh in names(results_summary[[resp]]$thresholds)) {
    # Thresholds
    write.csv(
      results_summary[[resp]]$thresholds[[thresh]], 
      sprintf("Output/Exp1_results_%s_thr_%s.csv", resp, tolower(thresh)),
      row.names = FALSE
    )
    
    # TLS metrics
    write.csv(
      results_summary[[resp]]$tls[[thresh]], 
      sprintf("Output/Exp1_results_%s_tls_%s.csv", resp, tolower(thresh)),
      row.names = FALSE
    )
  }
}

```

## Compile final results

```{r}

# Flatten TLS metrics
df_tls <- flatten_results_summary(results_summary)

# View the flattened data
tail(df_tls)

# Get summary statistics
summary_stats <- summarize_results(df_tls)
print(summary_stats)

# Compare valid vs invalid fits
validity <- compare_valid_invalid(df_tls)
print(validity)

# Flatten threshold data (T[n] across time)
df_thresholds <- flatten_threshold_data(results_summary)
head(df_thresholds)

# Export to CSV
write_csv(df_tls, "Output/Exp1_tls_metrics_flattened.csv")
write_csv(df_thresholds, "Output/Exp1_threshold_data_flattened.csv")

```

## Merge with metadata

```{r}

df_tls <- read.csv("Output/Exp1_tls_metrics_flattened.csv")
df_thresholds <- read.csv("Output/Exp1_threshold_data_flattened.csv")

str(df_tls)
str(dat)

# Get unique metadata per unique_sample (remove duplicate rows and irrelevant columns)
metadata <- dat %>%
  select(-X, -Temp, -Cell, -AOI) %>%
  distinct(unique_sample, .keep_all = TRUE)

# Create metadata keeping the Time variable
metadata_with_time <- dat %>%
  select(-X, -Temp, -Cell, -AOI) %>%
  distinct(unique_sample, Time, .keep_all = TRUE)

# Merge with df_tls
df_tls_merged <- df_tls %>%
  left_join(metadata, by = "unique_sample")

# Merge with df_thresholds
df_thresholds_merged <- df_thresholds %>%
  left_join(metadata_with_time, by = c("unique_sample", "Time"))

# View the result
head(df_tls_merged)
str(df_tls_merged)

head(df_thresholds_merged)
str(df_thresholds_merged)

df_thresholds_merged <- df_thresholds_merged %>%
  mutate(threshold = factor(threshold, levels = c("T10", "T0.3", "T50", "T90")))

df_tls_merged <- df_tls_merged %>%
  mutate(threshold = factor(threshold, levels = c("T10", "T0.3", "T50", "T90")))

```


## Fit statistical models

```{r}

# Model response of T90 io to recovery, duration, and species
df_t90_e1 <- df_thresholds_merged %>% filter(response_type=="fvfm_io" & threshold == "T90")
mod_t90_e1 <- lm(temperature ~ Recovery * Time * Species, data = df_t90_e1)
anova(mod_t90_e1)

# Model response of T50 io to recovery, duration, and species
df_t50_e1 <- df_thresholds_merged %>% filter(response_type=="fvfm_io" & threshold == "T50")
mod_t50_e1 <- lm(temperature ~ Recovery * Time * Species, data = df_t50_e1)
anova(mod_t50_e1)

# Model response of T10 io to recovery, duration, and species
df_t10_e1 <- df_thresholds_merged %>% filter(response_type=="fvfm_io" & threshold == "T10")
mod_t10_e1 <- lm(temperature ~ Recovery * Time * Species, data = df_t10_e1)
anova(mod_t10_e1)

# Effect of dark vs light * species on CTmax
df_tls_e1 <- df_tls_merged %>% filter(response_type=="fvfm_io")
mod_ctmax_e1 <- lm(ctmax_1h ~ Recovery * threshold + Species, data = df_tls_e1)
anova(mod_ctmax_e1)

mod_ctmax1m_e1 <- lm(ctmax_1m ~ Recovery * threshold + Species, data = df_tls_e1)
anova(mod_ctmax1m_e1)

mod_z_e1 <- lm(z_1h ~ Recovery * threshold + Species, data = df_tls_e1)
anova(mod_z_e1)

```

## Plot Exp 1 results

```{r}

# Create directory for output if it doesn't exist
if(!dir.exists("Plots")) {
  dir.create("Plots")
}

pred_data_individual <- df_tls_merged %>%
  filter((response_type == "fvfm_io" | response_type == "fvfm_absolute") & valid_fit == TRUE) %>%
  select(-Time) %>% 
  rowwise() %>%
  mutate(
    pred_df = list({
      time_seq <- seq(1, 120, length.out = 100)
      temp_pred <- ctmax_1m + (z_1m * log10(time_seq))
      data.frame(Time = time_seq, temperature_pred = temp_pred)
    })
  ) %>%
  unnest(pred_df)


# Then average predictions across samples
pred_data <- pred_data_individual %>%
  group_by(Species, Recovery, response_type, threshold, Time) %>%
  summarise(
    temperature_pred = mean(temperature_pred, na.rm = TRUE),
    .groups = "drop"
  )

sp_labels <- c(
  "Eucalyptus pauciflora" = "italic(Eucalyptus~pauciflora)",
  "Lomandra longifolia" = "italic(Lomandra~longifolia)",
  "Populus nigra" = "italic(Populus~nigra)",
  "Vicia faba" = "italic(Vicia~faba)"
)

sp_labeller <- function(variable, value) {
  return(sp_labels[value])
}


# Plot
ggplot(df_thresholds_merged %>%
         filter(response_type=="fvfm_io" | response_type=="fvfm_absolute"), 
       aes(x = Time, y = temperature)) +
  facet_grid(
    Species ~ Recovery,
    labeller = labeller(Species = as_labeller(sp_labels, label_parsed))
  ) +
  stat_summary(aes(colour = Recovery, fill = Recovery, shape = threshold),
               fun.data = "mean_se") +
  geom_line(data = pred_data, 
            aes(x = Time, y = temperature_pred, colour = Recovery, group = threshold),
            linewidth = 1, alpha = 0.5) +
  scale_shape_discrete(name = "Threshold", labels = c(bquote(italic(T)[10]),
                                                      bquote(italic(T)[0.3]),
                                                      bquote(italic(T)[50]),
                                                      bquote(italic(T)[90]))) +
  scale_colour_viridis_d(option = "inferno", end = 0.8, begin = 0.1, direction = 1,
                         labels = c("Dark", "Light"), name = "Recovery\ncondition") +
  scale_fill_viridis_d(option = "inferno", end = 0.8, begin = 0.1, direction = 1,
                         labels = c("Dark", "Light"), name = "Recovery\ncondition") +
  expand_limits(x = c(log10(1), log10(120)), y = c(30, 65)) +
  scale_y_continuous(limits = c(30, 65), breaks = seq(30, 65, by = 10)) +
  scale_x_log10(breaks = c(1,5,15,30,60,120),
                labels = c("1", "5", "15", "30", "60", "120")) +
  labs(x = "Exposure duration (min)", y = bquote(Heat~tolerance~(degree*C))) +
  theme_bw() + 
  theme(legend.position = "bottom")

ggsave("Plots/Fig_Exp1_1.pdf", width = 6, height = 8)

# TLS metrics plot and summary stats

temp_summary_by_threshold <- df_thresholds_merged %>%
  group_by(Species, Recovery, threshold) %>%
  summarise(
    n_obs = n(),
    min_temp = min(temperature, na.rm = TRUE),
    mean_temp = mean(temperature, na.rm = TRUE),
    max_temp = max(temperature, na.rm = TRUE),
    sd_temp = sd(temperature, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(threshold, Species, Recovery)

print(temp_summary_by_threshold, n = 32)


temp_summary_by_tls <- df_tls_merged %>%
  group_by(Species, Recovery, threshold) %>%
  summarise(
    n_obs = n(),
    mean_ctmax_1h = mean(ctmax_1h, na.rm = TRUE),
    sd_ctmax_1h = sd(ctmax_1h, na.rm = TRUE),
    mean_ctmax_1m = mean(ctmax_1m, na.rm = TRUE),
    sd_ctmax_1m = sd(ctmax_1m, na.rm = TRUE),
    mean_z_1h = mean(z_1h, na.rm = TRUE),
    sd_z_1h = sd(z_1h, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = starts_with("mean_"),
    names_to = "metric",
    values_to = "mean_value",
    names_prefix = "mean_"
  ) %>%
  # Add corresponding SD values
  mutate(
    sd_value = case_when(
      metric == "ctmax_1h" ~ sd_ctmax_1h,
      metric == "ctmax_1m" ~ sd_ctmax_1m,
      metric == "z_1h" ~ sd_z_1h
    )
  ) %>%
  select(-sd_ctmax_1h, -sd_ctmax_1m, -sd_z_1h) %>%
  arrange(threshold, Species, Recovery, metric)


metric_labels <- c(
  "ctmax_1h" = "italic(CT)[max_1h]",
  "ctmax_1m" = "italic(CT)[max_1m]",
  "z_1h" = "italic(z)"
)

threshold_labels <- c(
  "T10" = "italic(T)[10]",
  "T0.3" = "italic(T)[0.3]",
  "T50" = "italic(T)[50]",
  "T90" = "italic(T)[90]"
)

# Custom labeller function
threshold_labeller <- function(variable, value) {
  return(threshold_labels[value])
}

metric_labeller <- function(variable, value) {
  return(metric_labels[value])
}

ggplot(temp_summary_by_tls, 
       aes(x = Recovery, y = mean_value, color = Species, fill = Species)) +
  facet_grid(
    metric ~ threshold, scales = "free_y",
    labeller = labeller(threshold = as_labeller(threshold_labels, label_parsed),
                        metric = as_labeller(metric_labels, label_parsed))
  ) +
  geom_pointrange(aes(ymin = mean_value - sd_value, ymax = mean_value + sd_value), position = position_dodge(0.3)) +
  scale_colour_viridis_d(option = "mako", end = 0.8, begin = 0.1, direction = 1,
                         name = "Species") +
  labs(x = bquote(Recovery~condition), 
       y = bquote(Heat~tolerance~(degree*C))) +
  theme_bw() + theme(legend.position = "bottom", legend.text = element_text(face = "italic"))

ggsave("Plots/Fig_Exp1_2.pdf", width = 6, height = 5)

```